---
title: "FINARTS HW 3 — Market Model Regression"
subtitle: "Group 4 — Sison, Aaron Joshua E.; Opiana, Aimee Lorynne A.; Galedo, Enrique Lorenzo; Patajo, Juliana; Go, Keira; Cuenca, Raphael"
date: "June 2026"
format:
  pdf:
    toc: true
    number-sections: true
    pdf-engine: pdflatex
    pdf-version: "1.4"
    include-in-header:
      text: |
        \pdfobjcompresslevel=0
        \pdfminorversion=4
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

**Model:** $$R_i = \beta_0 + \beta_1 \cdot R_m + u$$

**Data files:**
- `Company_Returns.xlsx` — Daily returns of all active PSE-listed companies (Jan 1, 2023 – Jan 23, 2026), long format
- `PSEi_Return.xlsx` — Daily returns of the PSE index (Jan 1, 2023 – Jan 29, 2025)
- `Company Name.xlsx` — Company details (name, RIC, sector)

**Selected stocks (5 sectors):**

| Stock | RIC | Sector | Company |
|-------|-----|--------|---------|
| BPI | `BPI.PS` | Banks | Bank of the Philippine Islands |
| Century Pacific | `CNPF.PS` | Food Producers | Century Pacific Food |
| Meralco | `MER.PS` | Electricity | Manila Electric (Meralco) |
| ABS-CBN | `ABS.PS` | Media | ABS-CBN |
| PLDT | `TEL.PS` | Telecommunications | PLDT |

---

# Data Loading

```{r load-data}
library(dplyr)
library(tidyr)
library(readxl)
library(broom)

set.seed(42L)

company_returns <- read_excel("data/Company_Returns.xlsx")
psei_return    <- read_excel("data/PSEi_Return.xlsx")
company_info   <- read_excel("data/Company Name.xlsx")

cat("DATA LOADED\n")
cat("Company_Returns:", nrow(company_returns), "rows\n")
cat("PSEi_Return:",    nrow(psei_return),    "rows\n")
cat("Company Info:",   nrow(company_info),   "rows\n\n")
```

```{r stocks}
selected_rics <- c("BPI.PS", "CNPF.PS", "MER.PS", "ABS.PS", "TEL.PS")

stock_meta <- company_info |>
  filter(RIC %in% selected_rics) |>
  select(RIC, Name = Name, Sector, Full_Name = `Full Name`)

cat("SELECTED STOCKS\n")
print(stock_meta, row.names = FALSE)
cat("\n")
```

---

# PART 1: Regression Analysis

**Instruction:** Select 5 stocks from 5 different sectors. Run a regression of the returns of each stock against the market returns.

$$R_i = \beta_0 + \beta_1 \cdot R_m + u$$

---

## Question 1 (2 pts)

> Create a dataframe for your regressions. It must contain the 5 stock returns, PSEi return, and date.

```{r q1-dataframe}
psei <- psei_return |>
  mutate(Date = as.Date(Date)) |>
  rename(PSEi_Return = PSEi_Return) |>
  distinct(Date, .keep_all = TRUE)

stock_returns <- company_returns |>
  filter(RIC %in% selected_rics) |>
  rename(Return = Total_Return, Date = Date) |>
  mutate(Date = as.Date(Date))

reg_data_long <- stock_returns |>
  inner_join(psei, by = "Date")

cat("REGRESSION DATA (long format)\n")
cat("Rows:", nrow(reg_data_long), "\n")
cat("Date range:", format(min(reg_data_long$Date), "%Y-%m-%d"),
    "to", format(max(reg_data_long$Date), "%Y-%m-%d"), "\n\n")
```

```{r q1-wide}
reg_data_wide <- reg_data_long |>
  select(Date, RIC, Return, PSEi_Return) |>
  pivot_wider(
    id_cols     = c(Date, PSEi_Return),
    names_from  = RIC,
    values_from = Return,
    names_prefix = ""
  ) |>
  mutate(across(all_of(selected_rics), ~ as.numeric(.)))

stopifnot(all(sapply(reg_data_wide[selected_rics], is.numeric)))

cat("REGRESSION DATAFRAME (wide format)\n")
cat("Dimensions:", nrow(reg_data_wide), "rows x", ncol(reg_data_wide), "cols\n")
cat("Columns:", colnames(reg_data_wide), "\n")
cat("First 5 rows shown below.\n\n")
```

---

## Question 2 (2 pts each, 10 pts total)

> Run the regressions and report your estimated parameters. Follow this format when reporting:

> ```
> STOCK 1
> RIC: XX
> Company Name: XX
> Sector: XX
> Estimated parameters:
> B0: XX (indicate significance with asterisks)
> B1: XX (indicate significance with asterisks)
> ```

```{r q2-regressions}
stock_names <- list(
  "BPI.PS"  = "Bank of the Philippine Islands",
  "CNPF.PS" = "Century Pacific Food",
  "MER.PS"  = "Manila Electric (Meralco)",
  "ABS.PS"  = "ABS-CBN",
  "TEL.PS"  = "PLDT"
)

stock_sectors <- list(
  "BPI.PS"  = "Banks",
  "CNPF.PS" = "Food Producers",
  "MER.PS"  = "Electricity",
  "ABS.PS"  = "Media",
  "TEL.PS"  = "Telecommunications Service Providers"
)

sig_stars <- function(p_value) {
  if (is.na(p_value)) return("")
  if (p_value < 0.01) return("***")
  if (p_value < 0.05) return("**")
  if (p_value < 0.10) return("*")
  return("")
}

regression_results <- list()

cat("REGRESSION RESULTS\n")
cat("Model: R_i = B0 + B1 * R_m + u\n\n")

for (ric in selected_rics) {
  stock_data <- reg_data_long |> filter(RIC == ric)
  model <- lm(Return ~ PSEi_Return, data = stock_data)
  regression_results[[ric]] <- model

  coef_summary <- summary(model)$coefficients
  b0_est    <- coef_summary[1, 1]
  b1_est    <- coef_summary[2, 1]
  b0_pval   <- coef_summary[1, 4]
  b1_pval   <- coef_summary[2, 4]
  b0_sig    <- sig_stars(b0_pval)
  b1_sig    <- sig_stars(b1_pval)
  r_squared <- summary(model)$r.squared
  n_obs     <- nobs(model)

  cat("STOCK:", which(selected_rics == ric), "\n")
  cat("RIC:", ric, "\n")
  cat("Company Name:", stock_names[[ric]], "\n")
  cat("Sector:", stock_sectors[[ric]], "\n")
  cat("Estimated parameters:\n")
  cat("B0:", sprintf("%.6f%s", b0_est, b0_sig), "\n")
  cat("B1:", sprintf("%.6f%s", b1_est, b1_sig), "\n")
  cat("R-squared:", sprintf("%.4f", r_squared), "\n")
  cat("Observations:", n_obs, "\n\n")
}
```

### Detailed Output Per Stock

```{r q2-detail}
cat("DETAILED MODEL SUMMARIES\n\n")
for (ric in selected_rics) {
  cat("\n", ric, "\n")
  print(summary(regression_results[[ric]]))
  cat("\n")
}
```

---

## Question 3 (5 pts)

> Discuss your findings. Discuss the results for each stock, and compare the estimated parameters for each of the stocks.

**Per-stock analysis:**

Century Pacific Food (CNPF.PS) is the only stock with a negative beta value. A 1% increase in the PSEi leads to a 0.029% decrease in Century Pacific's return. This suggests a defensive characteristic, though the relationship is not statistically significant.

**Cross-stock comparison:**

Every stock's p-value is greater than 0.05, meaning the results are statistically insignificant. This indicates there is no correlation between our chosen stocks' returns and the PSEi returns from 2023 to 2025. Similarly, the R-squared results for each stock are close to zero, implying the PSEi returns only explain a tiny portion of the variance of price movements for each stock.

---

# PART 2: Fitted Values and Residuals

---

## Question 1 (5 pts)

> Using the regression dataframe that you used in Part 1, create columns where you compute the fitted values of the 5 stocks that you analyzed.

$$\hat{R}_{i,t} = \hat{\beta}_0 + \hat{\beta}_1 \cdot R_{m,t}$$

```{r p2q1-fitted}
reg_data_wide <- reg_data_wide |>
  mutate(
    BPI_Fitted   = fitted(regression_results[["BPI.PS"]]),
    CNPF_Fitted  = fitted(regression_results[["CNPF.PS"]]),
    MER_Fitted   = fitted(regression_results[["MER.PS"]]),
    ABS_Fitted   = fitted(regression_results[["ABS.PS"]]),
    TEL_Fitted   = fitted(regression_results[["TEL.PS"]])
  )

cat("FITTED VALUES ADDED\n")
cat("New columns:", grep("Fitted", colnames(reg_data_wide), value = TRUE), "\n\n")
```

---

## Question 2 (5 pts)

> Still in the same dataframe, compute the residuals for each stock.

$$\hat{u}_{i,t} = R_{i,t} - \hat{R}_{i,t}$$

```{r p2q2-residuals}
reg_data_wide <- reg_data_wide |>
  mutate(
    BPI_Residual   = BPI.PS   - BPI_Fitted,
    CNPF_Residual  = CNPF.PS  - CNPF_Fitted,
    MER_Residual   = MER.PS   - MER_Fitted,
    ABS_Residual   = ABS.PS   - ABS_Fitted,
    TEL_Residual   = TEL.PS   - TEL_Fitted
  )

cat("RESIDUALS ADDED\n")
cat("New columns:", grep("Residual", colnames(reg_data_wide), value = TRUE), "\n\n")
```

---

## Question 3 (3 pts)

> What did you notice about the sum of the residuals for each stock?

```{r p2q3-sum-residuals}
residual_sums <- reg_data_wide |>
  summarise(
    BPI_Sum   = sum(BPI_Residual,   na.rm = TRUE),
    CNPF_Sum  = sum(CNPF_Residual,  na.rm = TRUE),
    MER_Sum   = sum(MER_Residual,   na.rm = TRUE),
    ABS_Sum   = sum(ABS_Residual,   na.rm = TRUE),
    TEL_Sum   = sum(TEL_Residual,   na.rm = TRUE)
  )

cat("SUM OF RESIDUALS\n")
knitr::kable(residual_sums,
             col.names = c("BPI", "CNPF", "MER", "ABS", "TEL"),
             digits = 6,
             booktabs = TRUE,
             caption = "Sum of Residuals by Stock")
cat("\n")

for (ric_name in c("BPI", "CNPF", "MER", "ABS", "TEL")) {
  col_name <- paste0(ric_name, "_Sum")
  sum_val <- residual_sums[[col_name]]
  cat(ric_name, ": ", sprintf("%.6f", sum_val), "\n", sep = "")
}
cat("\n")
```

The sum of residuals for every stock is close to zero as a result of their respective residuals cancelling out.

---

# Data Export

```{r export, include=FALSE}
output_csv <- "H3_Group4_Regression_Output.csv"
write.csv(reg_data_wide, file = output_csv, row.names = FALSE)
cat("CSV EXPORTED\n")
cat("File:", output_csv, "\n")
cat("Dimensions:", nrow(reg_data_wide), "rows x", ncol(reg_data_wide), "cols\n")
cat("Columns:", paste(colnames(reg_data_wide), collapse = ", "), "\n\n")
```

```
Date, PSEi_Return, BPI.PS, MER.PS, TEL.PS, CNPF.PS, ABS.PS,
BPI_Fitted, CNPF_Fitted, MER_Fitted, ABS_Fitted, TEL_Fitted,
BPI_Residual, CNPF_Residual, MER_Residual, ABS_Residual, TEL_Residual
```

---

# Summary of Submitted Files

| # | File | Description |
|---|------|-------------|
| 1 | `H3_Group4_Regression_Output.csv` | Final dataframe (727 rows × 17 cols) |
| 2 | `H3_Analysis.R` | Standalone R analysis script |
| 3 | `H3.pdf` | This document — PDF with answers |
