---
title: "FINARTS HW 3 — Market Model Regression"
author: "`Group 4 \\\\ Cuenca, Raphael \\\\ Galedo, Enrique Lorenzo \\\\ Go, Keira \\\\ Opiana, Aimee Lorynne A. \\\\ Patajo, Juliana \\\\ Sison, Aaron Joshua E.`{=latex}"
date: "June 16, 2026"
format:
  pdf:
    toc: true
    number-sections: true
    pdf-engine: pdflatex
    colorlinks: true
    geometry: "margin=1in"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
options(scipen = 999) # Disable scientific notation for clearer output
knitr::read_chunk("H3_Analysis.R")
```

\newpage

# Overview

**Model:** $$R_i = \beta_0 + \beta_1 \cdot R_m + u$$

**Selected stocks (5 sectors):**

| Stock | RIC | Sector | Company |
|-------|-----|--------|---------|
| BPI | `BPI.PS` | Banks | Bank of the Philippine Islands |
| Century Pacific | `CNPF.PS` | Food Producers | Century Pacific Food |
| Meralco | `MER.PS` | Electricity | Manila Electric (Meralco) |
| ABS-CBN | `ABS.PS` | Media | ABS-CBN |
| PLDT | `TEL.PS` | Telecommunications | PLDT |

---

# Data Loading

```{r load-data}
```

---

# PART 1: Regression Analysis

**Instruction:** Select 5 stocks from 5 different sectors. Run a regression of the returns of each stock against the market returns.

$$R_i = \beta_0 + \beta_1 \cdot R_m + u$$

## Question 1 (2 pts)

> Create a dataframe for your regressions. It must contain the 5 stock returns, PSEi return, and date.

```{r q1-dataframe}
```

---

## Question 2 (2 pts each, 10 pts total)

> Run the regressions and report your estimated parameters. 

```{r q2-regressions}
```

*(Note on significance: *** $p<0.01$, ** $p<0.05$, * $p<0.10$)*

---

## Question 3 (5 pts)

> Discuss your findings. Discuss the results for each stock, and compare the estimated parameters for each of the stocks.

The empirical estimation of the market model across the five selected equities reveals a pervasive lack of statistical significance at the standard 5% threshold ($p > 0.05$ for all $\beta_1$ coefficients). For the timeframe evaluated (2023–2025), the systematic market movements dictated by the PSEi failed to meaningfully explain the daily variance of these specific stocks, a finding corroborated by near-zero $R^2$ values across the panel. Consequently, the daily returns of these selected equities appear to be driven predominantly by idiosyncratic, company-specific factors rather than broader macroeconomic market trends.

Despite the overarching statistical insignificance, examining the individual slope coefficients provides insight into the directional sensitivities of the equities. PLDT (TEL.PS) registers the highest market sensitivity among the cohort with a beta of $0.0659$, emerging as the only stock to approach marginal significance at the 10% level ($p = 0.083$). This suggests it tracks the broader index slightly more closely than its peers. Similarly, Manila Electric (MER.PS) yields a positive beta ($0.0338$) coupled with a marginally significant positive alpha ($\beta_0 = 0.1210$, $p < 0.10$), hinting at slight positive excess returns generated independently of the market. Both the Bank of the Philippine Islands (BPI.PS) and ABS-CBN (ABS.PS) exhibit very weak positive betas ($0.0145$ and $0.0240$, respectively) alongside high p-values, confirming that market fluctuations barely permeate their return profiles.

In contrast to the rest of the cohort, Century Pacific Food (CNPF.PS) stands out as the only stock yielding a negative beta value. Specifically, the estimated parameters indicate that a 1% increase in the PSEi leads to a 0.029% decrease in Century Pacific's return. This inverse dynamic suggests a defensive, counter-cyclical characteristic inherent to the food producer, even though the relationship itself ultimately fails to achieve statistical significance ($p = 0.539$).

---

# PART 2: Fitted Values and Residuals

## Question 1 (5 pts)

> Compute the fitted values of the 5 stocks that you analyzed.

$$\hat{R}_{i,t} = \hat{\beta}_0 + \hat{\beta}_1 \cdot R_{m,t}$$

```{r p2q1-fitted}
```

## Question 2 (5 pts)

> Compute the residuals for each stock.

$$\hat{u}_{i,t} = R_{i,t} - \hat{R}_{i,t}$$

```{r p2q2-residuals}
```

## Question 3 (3 pts)

> What did you notice about the sum of the residuals for each stock?

```{r p2q3-sum-residuals}
```

The sum of residuals for every stock is practically zero. This is a fundamental mathematical property of Ordinary Least Squares regression when an intercept is included. The positive and negative errors perfectly cancel each other out, leaving only a microscopic remainder due to computer precision limits. In finance, these residuals represent the specific returns of a company that are independent of the broader market. A zero sum confirms that our market model is unbiased over the evaluation period. The model intercept fully captures the average excess return of the stock. Therefore, any firm specific surprises naturally average out to zero over time. This proves that the remaining unexplained returns represent purely diversifiable risk.



# Data Export

```{r export}
```

The final dataframe has been successfully exported to `H3_Group4_Regression_Output.csv` containing **`r nrow(reg_data_wide)` rows** and **`r ncol(reg_data_wide)` columns** (Date, PSEi_Return, 5 stock returns, 5 fitted values, 5 residuals).
